Runaway and Hypervelocity Stars from Compact Object Encounters in Globular Clusters

### 2023-08-25 src/tex/ms.tex
Updated links, removed ApJ formatting.

We implement additional features into \fewbody\ pertinent to the study of compact object dynamics.
We add additional parameters to track stellar object type, distinguishing between stars and compact objects.%, using the numbering convention of \citet{2002MNRAS.329..897H}; similarly, we implement the collision matrix presented in Table 2 of the same paper to determine the type of any post-merger products.
We use these object types to decide whether to use \fewbody's default sticky-sphere collision criterion - which triggers a collision when the separation of the objects' centers is less than the sum of their radii - in the case of stellar interactions, and a tidal disruption criterion in the case of encounters involving NSs and BHs, which multiplies the threshold separation by a factor of $(m_{\rm CO} / m_{*})^{1/3}$ \citep[as was done in][]{Kremer2019}.
To the utility of this collision metric, we conservatively assign BHs radii of 5 times their Schwarzschild radii to group the more extreme encounters with mergers, both of which require more careful treatment to accurately represent.

To establish realistic binary-single encounter populations for clusters of various parameters, we take the 148 models from the \CMCcat\ and extract the initial conditions of all strong binary-single encounters that involve at least one luminous object (star, WD, or NS; to limit the focus to encounters that can produce observable ejecta) and one CO.
Over the entire catalog, our encounter sample makes up about half of all strong encounters that occur in the models, with each model contributing a few ten thousand BSCO encounters on average.
We then realize each encounter in isolation with our modified \fewbody\, computing 10 realizations of each encounter while re-drawing the randomly selected parameters (e.g.~binary phase) to obtain a better statistical representation of the binary-single encounter population.

@@

The initial conditions used for these encounters are calculated in the center-of-mass rest frame of the encounter, i.e. they do not contain information about the center-of-mass velocity of the encounter in the frame of the GC model.
We do not attempt to correct for this, which leads to a minor underestimation of final velocities for encountering objects and subsequently ejection rates; we justify this in Appendix \ref{app:restframe}.

@@

First core collapse is stopped by the formation and hardening of binaries in the core (which at this point is largely composed of BHs), and the liberated energy supports the universal expansion of the GC for a time on the order of Gyr \citep{2012MNRAS.420..309B}.
The BH population in the core is gradually depleted as strong encounters accelerate their participants beyond the local escape speed, until no more than a handful of BHs remain in the cluster \citep{2013MNRAS.432.2779B, 2020IAUS..351..357K}.
The loss of the dynamical heat source causes the core to contract again, as the system adjusts to being supported by WDs, NSs and stars in the same mass regime.
Unlike the first collapse, this second, permanent core collapse is observable in that the surface brightness profile of the GC becomes cuspy to the limit of the cluster center (and is what observers would classically call a ``core-collapsed cluster'').
We follow \citet{2018ApJ...855L..15K} in pairing these demographic and observable features, marking the time of second core collapse in our models as the same time the number of BHs drops below 10.

@@

Figure \ref{fig:cmc_single_clusters} shows the ejection velocity $v_{\rm out}$ for every MS star ejected through a BSCO encounter in these models, plotted at the time of encounter from initialization $t$; \textbf{generally, these ejecta account for $\sim$1-6\% of all stellar escapers from a model over its lifetime}.
What is immediately apparent from these single cluster models is a strong correlation between BSCO encounters and the evolution of the cluster core: whenever a cluster undergoes a core collapse, the heightened densities lead to predictably high encounter rates.

@@

An interplay between stellar evolution and cluster dynamics is revealed in the kinds of BSCO encounters that occur at certain times.
The most massive stars in a cluster are both the first to form binaries and the first to evolve into COs.
hence, encounters between a mixed binary (composed of a MS star and a CO) and a single MS star \textbf{have the first opportunity to contribute significantly to the overall BSCO encounter population; in the densest models, this primacy leads to a majority of BSCO encounters being of this type over the 14 Gyr evolution.
Multiple-CO encounters first become major players at first core collapse after CO formation, and can take over as the majority in the case of models with weak or nonexistent pre-core collapse BSCO encounter phases.}
There is a stark difference in the ejection velocities produced by these multiple-CO encounters in comparison to the previous mixed binary-single MS star encounters, and the fastest ejections over the cluster's lifetime are produced in the first tens of Myr after the associated first core collapse.
This regime lasts for a few to tens of Gyr, during which the COs in the core (predominantly BHs) preferentially form CO-CO binaries, as evidenced in the drop of mixed binary-single CO encounters during this phase.

@@

The dominance of the core in overall BSCO encounter production is even more clear when considering the radial localization of the encounters.
\CMC\ does not record the radius where each strong encounter occurred, but it does record the local escape velocity.
For comparison, the escape velocity from the center of the model can be computed from the central and tidal boundary potentials, which are recorded throughout the evolution of the cluster.
\textbf{
    Figure \ref{fig:cmc_single_clusters_vesc} plots these data for the first sample cluster, where the distribution of encounter escape velocities is strikingly correlated to the central escape velocity, indicating the degree to which the core dominates these dynamics.
    Throughout the evolution of the model, most BSCO ejections occur at or near the central escape velocity, and this feature is common to the counterpart plots for the other sample models.
    There is a small fraction of points that lie above the core velocity line, which we attribute to the difference in methods employed to calculate escape velocities in the ejection and core contexts.
}

\begin{figure}
    \script{cmc_single_clusters_vesc.py}
    \centering
    \includegraphics[width=0.49\textwidth]{figures/cmc_single_clusters_vesc.pdf}
    \caption{
        \textbf{
        The local escape velocities $v_{\rm esc}$ at the locations of all BSCO encounters generated from the first sample \CMC\ model (colored points), plotted alongside the central escape velocity of the model (navy).
        The top plot shows the data in units of ${\rm km~s^{-1}}$, while the lower plot shows the same data normalized to the continuum of the core escape velocity (navy line in the upper plot).
        }
        See Figure \ref{fig:cmc_single_clusters} for an explanation of the color scheme.
    }
    \label{fig:cmc_single_clusters_vesc}
\end{figure}

@@

\begin{figure}
    \script{mwej_rates.py}
    \begin{centering}
        \includegraphics[width=0.99\linewidth]{figures/mwej_rates.pdf}
        \caption{
            The evolution of the ejection rate for different components of our synthetic population.
            The blue (red) dashed line shows the estimated from observations for runaways from the galactic disc (HVSs from the galactic center), as reported by \citet{2015ARA&A..53...15B}.
            \textbf{The threshold times by which 50\% and 90\% of the runaway ejections have occured are indicated along the bottom of the figure.}
        }
        \label{fig:mwej_rates}
    \end{centering}
\end{figure}

@@

Finally, Figure \ref{fig:mwej_rates} shows the HVS/runaway rates for our synthetic population, along with observational rate estimates from \citet{2015ARA&A..53...15B}.
\textbf{We note that 50(90)\% of ejections occur before our synthetic population is \variable{output/mwej_rates.txt}old, and that naturally the earliest ejections are most dependent on any time variability in the MW potential.}
At these earlier times, BSCO ejecta could have made as much as 20\% (1\%) of all runaways (HVSs), during the times when GCs underwent their first core collapses.\footnote{Here, the cutoff of this phase is fairly abrupt because we constrain our models to present-day ages of 10-13.5 Gyr (\S\ref{subsec:pairing}).  The physical rate will naturally be characterized by the evolution of the GC formation rate in the past.}
Afterwards, the ejection rates decay to the present day, where they are closer to 1.5\% (0.0001\%) of the observational values.
It is important to note nonetheless that GCs are able to produce HVSs with speeds high enough to make them comparable to those accelerated by mechanisms in the galactic center.

@@

This work was prepared in part with the reproducibility software \href{https://github.com/showyourwork/showyourwork}{\showyourwork}\citep{2021arXiv211006271L}, which leverages continuous integration to transparently connect the paper to the publicly-available dataset and relevant code.
The GitHub repository for this project can be found at \url{https://github.com/tomas-cabrera/hvss-bsco}; the icons next to paper content link to the particular scripts used to generate that content.
The dataset is stored at \url{https://doi.org/10.5281/zenodo.7599870}, which also includes copies of the simulation and figure scripts from the GitHub for insurance of future access.

@@

We include with this publication phase credible regions in phase space for ejecta populations from the MWGCs considered, which may be accessed at \url{https://doi.org/10.5281/zenodo.7599870}.
We intend these constructions to be used as secondary evidence for cluster membership of runaway stars, and to encourage further studies into understanding the respective phase space distributions.
The 2D position and proper motion distributions are separated into two different files, the conventions of which are described below.

The distribution of ejecta on the sky is quantified by discretizing the sphere with an order 4 nested HEALPix map \citep{2005ApJ...622..759G}, binning the sky into 3072 equal-area tiles.
We choose this resolution to enable identification of interesting credible regions while minimizing the apparent effect of isolated points whose exact location is dependent on the RNG seeds used.
We create the respective histogram by counting the number of synthetic ejecta that are found in each tile in the present day, and normalize by the total number of ejecta for the GC.
The resulting probability histogram is stored in the first column (\texttt{PROBS}) of the \texttt{hp\_probs.fits} file for each GC; the \texttt{NEJECT} field in the header of the same file contains the number of ejections for the GC.
We calculate our credible regions by cumulatively adding the highest probability bins until the target percentage is reached.
The last two columns (\texttt{CR50} and \texttt{CR90}) are boolean masks of the same convection as the HEALPix probability histogram corresponding to the credible regions (50\% and 90\%, respectively), where bins with entries of 1 are included in the region.
Figure \ref{fig:check_credible_regions_x} shows the histogram and 50\% and 90\% credible regions for NGC 104, as an example (the script used to generate this plot is \texttt{check\_credible\_regions.py} in the GitHub repository; click on the icon by the caption to be taken directly to the file).

>>>>

We implement additional features into \fewbody\ pertinent to the study of compact object dynamics.
We add additional parameters to track stellar object type, distinguishing between stars and compact objects.
We use these object types to decide whether to use \fewbody's default sticky-sphere collision criterion (which triggers a collision when the separation of the objects' centers is less than the sum of their radii) in the case of stellar interactions, or a tidal disruption criterion in the case of encounters involving NSs and BHs, which multiplies the threshold separation by a factor of $(m_{\rm CO} / m_{*})^{1/3}$ \citep[as was done in][]{Kremer2019}.
To the utility of this collision metric, we conservatively assign BHs radii of 5 times their Schwarzschild radii to group the more extreme encounters with mergers, both of which require more careful treatment to accurately compute.

To establish realistic binary-single encounter populations for clusters of various parameters, we take the 148 models from the \CMCcat\ and extract the initial conditions of all strong binary-single encounters that involve at least one luminous object (star, WD, or NS; to limit the focus to encounters that can produce observable ejecta) and one CO.
Over the entire catalog, our encounter sample makes up about half of all strong encounters that occur in the models, with each model contributing a few ten thousand BSCO encounters on average.
We realize each encounter in isolation with our modified \fewbody\, computing 10 realizations of each encounter while re-drawing the randomly selected parameters (e.g.~binary phase) to obtain a better statistical representation of the binary-single encounter population.

@@

The initial conditions used for these encounters are calculated in the center-of-mass rest frame of the encounter, i.e. they do not contain information about the center-of-mass velocity of the encounter in the frame of the GC model.
We do not attempt to correct for this, which leads to a minor underestimation of final velocities for encountering objects and subsequently ejection rates (we justify this in Appendix \ref{app:restframe}).

@@

First core collapse is stopped by the formation and hardening of binaries in the core (which at this point is largely composed of BHs), and the liberated energy supports the universal expansion of the GC for a time on the order of Gyr \citep{2012MNRAS.420..309B}.
The BH population in the core is gradually depleted as strong encounters accelerate their participants beyond the local escape speed, until no more than a handful of BHs remain in the cluster \citep{2013MNRAS.432.2779B, 2020IAUS..351..357K}.
The loss of the dynamical heat source causes the core to contract again, as the system adjusts to being supported by WDs, NSs and stars in the same mass regime.
Unlike the first collapse, this second, permanent core collapse is observable in that the surface brightness profile of the GC becomes cuspy to the limit of the cluster center (and is what observers would classically call a ``core-collapsed cluster'').
We follow \citet{2018ApJ...855L..15K} in pairing these demographic and observable features, marking the time of second core collapse in our models as the time when the number of BHs drops below 10.

@@

Figure \ref{fig:cmc_single_clusters} shows the ejection velocity $v_{\rm out}$ for every MS star ejected through a BSCO encounter in these models, plotted at the time of encounter from initialization $t$; generally, these ejecta account for $\sim$1-6\% of all stellar escapers from a model over its lifetime.
What is immediately apparent from these single cluster models is a strong correlation between BSCO encounters and the evolution of the cluster core: whenever a cluster undergoes a core collapse, the heightened densities lead to predictably high encounter rates.

@@

An interplay between stellar evolution and cluster dynamics is revealed in the kinds of BSCO encounters that occur at certain times.
The most massive stars in a cluster are both the first to form binaries and the first to evolve into COs.
Hence, encounters between a mixed binary (composed of a MS star and a CO) and a single MS star have the first opportunity to contribute significantly to the overall BSCO encounter population; in the densest models, this primacy leads to a majority of BSCO encounters being of this type over the 14 Gyr evolution.
Multiple-CO encounters first become major players at first core collapse after CO formation, and can take over as the majority in the case of models with weak or nonexistent pre-core collapse BSCO encounter phases.
There is a stark difference in the ejection velocities produced by these multiple-CO encounters in comparison to the previous mixed binary-single MS star encounters, and the fastest ejections over the cluster's lifetime are produced in the first tens of Myr after the associated first core collapse.
This regime lasts for a few to tens of Gyr, during which the COs in the core (predominantly BHs) preferentially form CO-CO binaries, as evidenced in the drop of mixed binary-single CO encounters during this phase.

@@

The dominance of the core in overall BSCO encounter production is even more clear when considering the radial localization of the encounters.
\CMC\ does not record the radius where each strong encounter occurred, but it does record the local escape velocity.
For comparison, the escape velocity from the center of the model can be computed from the central and tidal boundary potentials, which are recorded throughout the evolution of the cluster.
Figure \ref{fig:cmc_single_clusters_vesc} plots these data for the first sample cluster, where the distribution of encounter escape velocities is strikingly correlated to the central escape velocity, indicating the degree to which the core dominates these dynamics.
Throughout the evolution of the model, most BSCO ejections occur at or near the central escape velocity (the same plots for the other sample models also display this feature).
There is a small fraction of points that lie above the core velocity line, which we attribute to the difference in methods employed to calculate escape velocities in the ejection and core contexts.

\begin{figure}
    \script{cmc_single_clusters_vesc.py}
    \centering
    \includegraphics[width=0.49\textwidth]{figures/cmc_single_clusters_vesc.pdf}
    \caption{
        The local escape velocities $v_{\rm esc}$ at the locations of all BSCO encounters generated from the first sample \CMC\ model (colored points), plotted alongside the central escape velocity of the model (dark blue).
        The top plot shows the data in units of ${\rm km~s^{-1}}$, while the lower plot shows the same data normalized to the continuum of the core escape velocity (dark blue line in the upper plot).
        See Figure \ref{fig:cmc_single_clusters} for an explanation of the color scheme.
    }
    \label{fig:cmc_single_clusters_vesc}
\end{figure}

@@

\begin{figure}
    \script{mwej_rates.py}
    \begin{centering}
        \includegraphics[width=0.99\linewidth]{figures/mwej_rates.pdf}
        \caption{
            The evolution of the ejection rate for different components of our synthetic population.
            The blue (red) dashed line shows the estimated from observations for runaways from the galactic disc (HVSs from the galactic center), as reported by \citet{2015ARA&A..53...15B}.
            The threshold times by which 50\% and 90\% of the runaway ejections have occured are indicated along the bottom of the figure.
        }
        \label{fig:mwej_rates}
    \end{centering}
\end{figure}

@@

Finally, Figure \ref{fig:mwej_rates} shows the HVS/runaway rates for our synthetic population, along with observational rate estimates from \citet{2015ARA&A..53...15B}.
We note that 50(90)\% of ejections occur before our synthetic population is 3.21(8.59) Gyr old, and that naturally the earliest ejections are most dependent on any time variability in the MW potential.
At these earlier times, BSCO ejecta could have made as much as 20\% (1\%) of all runaways (HVSs), during the times when GCs underwent their first core collapses.\footnote{Here, the cutoff of this phase is fairly abrupt because we constrain our models to present-day ages of 10-13.5 Gyr (\S\ref{subsec:pairing}).  The physical rate will naturally be characterized by the evolution of the GC formation rate in the past.}
Afterwards, the ejection rates decay to the present day, where they are closer to 1.5\% (0.0001\%) of the observational values.
It is important to note nonetheless that GCs are able to produce HVSs with speeds high enough to make them comparable to those accelerated by mechanisms in the galactic center.

@@

This work was prepared in part with the reproducibility software \href{https://github.com/showyourwork/showyourwork}{showyourwork} \citep{2021arXiv211006271L}, which leverages continuous integration to transparently connect the paper to the publicly-available dataset and relevant code.
The GitHub repository for this project can be found at \url{https://github.com/tomas-cabrera/hvss-bsco}; the icons next to paper content link to the particular scripts used to generate that content.
The dataset is stored at \url{https://doi.org/10.5281/zenodo.7599870}, which also includes copies of the simulation and figure scripts from the GitHub for insurance of future access.

@@

We include with this publication phase credible regions in phase space for ejecta populations from the MWGCs considered, which may be accessed at \url{https://doi.org/10.5281/zenodo.7852916}.
We intend these constructions to be used as secondary evidence for cluster membership of runaway stars, and to encourage further studies into understanding the respective phase space distributions.
The 2D position and proper motion distributions are separated into two different files, the conventions of which are described below.

The distribution of ejecta on the sky is quantified by discretizing the sphere with an order 4 nested HEALPix map \citep{2005ApJ...622..759G}, binning the sky into 3072 equal-area tiles.
We choose this resolution to enable identification of interesting credible regions while minimizing the apparent effect of isolated points whose exact location is dependent on the RNG seeds used.
We create the respective histogram by counting the number of synthetic ejecta that are found in each tile in the present day, and normalize by the total number of ejecta for the GC.
The resulting probability histogram is stored in the first column (\texttt{PROBS}) of the \texttt{hp\_probs.fits} file for each GC; the \texttt{NEJECT} field in the header of the same file contains the number of ejections for the GC.
We calculate our credible regions by cumulatively adding the highest probability bins until the target percentage is reached.
The last two columns (\texttt{CR50} and \texttt{CR90}) are boolean masks of the same convection as the HEALPix probability histogram corresponding to the credible regions (50\% and 90\%, respectively), where bins with entries of 1 are included in the region.
Figure \ref{fig:check_credible_regions_x} shows the histogram and 50\% and 90\% credible regions for NGC 104, as an example (the script used to generate this plot is \texttt{check\_credible\_regions.py}, which may be found in the GitHub/Zenodo repositories).
